Modelling state-switching functional data with hidden Markov models: supplement B (simulation study)

### 2021-06-24 Report/Sections/results.tex
Update Report

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}. 
Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of the animal.

@@

The primary goal of this case study is to jointly estimate the dive types and (more importantly) the subdive behaviours of this killer whale, so we only consider models with a fine-scale component to describe the kinematic data. Defining a suitable hierarchical model involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.

@@

We then select the specific emission distributions of $\Z_{t,\tilde t^*}$ for all dive types and subdive states. First, we assume that $\Ztwo_{t,\tilde t^*}$ and all three components of $\Zone_{t,\tilde t^*}$ are independent of one another when conditioned on the dive types and subdive states. To reduce model complexity, we also assume that the three sets of fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$. This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\{\Zone_{t,1},\ldots,\Zone_{t,T^*_t}\}$ for a particular dive $t$. We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X_{t,1},\ldots,\tilde X_{t,\tilde T^*_t}$. Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter. Thus the distribution of $\Zone_{t,\tilde t^*}$ given $X^*_{t,\tilde t^*} = i^*$ has parameters $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
To specify the emission distribution of $\Ztwo_{t,\tilde t^*}$, we assume that given $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ follows a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. In addition, $\Ztwo_{t,1},\ldots,\Ztwo_{t,T^*_t}$ are assumed to be independent of one another given the subdive state sequence $\big\{\tilde X_{t,1}, \ldots, \tilde X_{t,T_t^*}\big\}$. We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.

@@

%To illustrate an application of this method and compare the candidate models, 
We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{scipy} package in Python \citep{Virtanen:2019}. Each model is fit 100 times using random initializations, and
%Let $\mu_0,\sigma_0$, and $\phi_0$ represent the initial mean, standard deviation, and auto-correlation parameters for a given observation. If the observation is strictly non-negative, then we set $\log(\mu_0) \sim \mathcal{N}(\log(\bar Y),s_{\log})$, $\log(\sigma) \sim \mathcal{N}(s_{\log},1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean of the observations and $s_{\log}$ represents the sample standard deviation the logarithm of the observations values. If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values. 
we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines (see Section 3.4 of \citet{Zucchini:2016} for details). Plots of the likelihood surface near the maximum likelihood parameter estimates are shown in Section 2 of Supplement A.

@@

The means of ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ associated with each subdive state are separated by an order of magnitude (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest standard deviation corresponding to $\Zone_{t,\tilde t^*}$. It also has the highest auto-correlation in $\Zone_{t,\tilde t^*}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ one order of magnitude higher than subdive state 1 and its acceleration has about twice the standard deviation compared to subdive state 1. The auto-correlation of acceleration is also slightly lower than subdive state 1. We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states, and the auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

@@

\[\hat \delta^{*(1)} = \begin{pmatrix} 0.202 & 0.649 & 0.149 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.531 & 0.405 & 0.064 \end{pmatrix}\]
%
for dive types 1 and 2.
In summary, about 86\% of dives are short dives of type 1, and the whale performs an average of 6.54 short type 1 dives before switching to dive type 2 and an average of 1.09 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
Further, this killer whale is in the less active subdive state 1 only 20\% of the time during a dive of type 1 compared to 53\% of the time during a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 4 of of supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.

@@

The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize 
any given behaviour as subdive state 3
%the behaviour at the beginning and end of dives as subdive state 3 
(see Figures 11 and 12 of supplement A). In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are %approximately $50$ to $100$ percent 
significantly larger than those of the CarHHMM-DFT. The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of of supplement A). This suggests that including auto-correlation in the model significantly affects parameter estimates. %Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of of supplement A). These findings suggest
and that the HHMM-DFT is a significantly worse fit to these data than the full CarHHMM-DFT.

@@

For a more complete set of results for each of the candidate models, see Supplement A.

>>>>

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For killer whales, we are interested in categorizing different diving behaviours and identifying potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}. 
Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of the animal.

@@

The primary goal of this case study is to jointly estimate the dive types and subdive behaviours of this killer whale, so we only consider models with both a coarse-scale and fine-scale component to describe the kinematic data. Defining a suitable hierarchical model involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.

@@

We then select the specific emission distributions of $\Z_{t,\tilde t^*}$ for all dive types and subdive states. First, we assume that $\Ztwo_{t,\tilde t^*}$ and all three components of $\Zone_{t,\tilde t^*}$ are independent of one another when conditioned on the dive types and subdive states. To reduce model complexity, we also assume that the three sets of fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$. This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\{\Zone_{t,1},\ldots,\Zone_{t,T^*_t}\}$ for a particular dive $t$. We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X^*_t = \{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T^*_t}\}$. Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter. Thus the distribution of $\Zone_{t,\tilde t^*}$ given $X^*_{t,\tilde t^*} = i^*$ has parameters $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
To specify the emission distribution of $\Ztwo_{t,\tilde t^*}$, we assume that given $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ follows a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. In addition, $\Ztwo_{t,1},\ldots,\Ztwo_{t,T^*_t}$ are assumed to be independent of one another given the subdive state sequence $\big\{\tilde X_{t,1}, \ldots, \tilde X_{t,T_t^*}\big\}$. We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.

@@

%To illustrate an application of this method and compare the candidate models, 
We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{scipy} package in Python \citep{Virtanen:2019}. Each model is fit 100 times using random initializations, and
%Let $\mu_0,\sigma_0$, and $\phi_0$ represent the initial mean, standard deviation, and auto-correlation parameters for a given observation. If the observation is strictly non-negative, then we set $\log(\mu_0) \sim \mathcal{N}(\log(\bar Y),s_{\log})$, $\log(\sigma) \sim \mathcal{N}(s_{\log},1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean of the observations and $s_{\log}$ represents the sample standard deviation the logarithm of the observations values. If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values. 
we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines (see Section 3.4 of \citet{Zucchini:2016} for details). Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory. Most, but not all, initializations converge to the same parameter estimates, highlighting the need to perform multiple optimizations with a variety of initial guesses. Plots of the likelihood surface near the maximum likelihood estimates are shown in Section 2 of supplement A.

@@

The means of ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ associated with each subdive state are separated by an order of magnitude (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest standard deviation corresponding to $\Zone_{t,\tilde t^*}$. It also has the highest auto-correlation in $\Zone_{t,\tilde t^*}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ one order of magnitude higher than subdive state 1 and its acceleration has about twice the standard deviation compared to subdive state 1. The auto-correlation of acceleration is also slightly lower than subdive state 1. We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states, and the auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower. This corresponds to vigorous swimming activity, especially as the killer whale ends a dive (see Figure \ref{fig:labeled_dives}).

@@

\[\hat \delta^{*(1)} = \begin{pmatrix} 0.202 & 0.649 & 0.149 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.531 & 0.405 & 0.064 \end{pmatrix}\]
%
for dive types 1 and 2.
In summary, about 86\% of dives are short dives of type 1, and the whale performs an average of 6.54 short type 1 dives before switching to dive type 2 and an average of 1.09 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
Further, this killer whale is in the less active subdive state 1 only 20\% of the time during a dive of type 1 compared to 53\% of the time during a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 4 of supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.

@@

The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize 
any given behaviour as subdive state 3
%the behaviour at the beginning and end of dives as subdive state 3 
(see Figures 11 and 12 of supplement A). In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are %approximately $50$ to $100$ percent 
significantly larger than those of the CarHHMM-DFT. The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of of supplement A). This suggests %that including auto-correlation in the model significantly affects parameter estimates, %Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of of supplement A). These findings suggest 
that the HHMM-DFT overlooks some auto-correlation in the data, and that the HHMM-DFT may be overconfident in its parameters estimates compared to the full CarHHMM-DFT.

@@

For a more complete set of results for each of the candidate models, see supplement A.

### 2021-06-24 Report/Sections/simulation.tex
Update Report

\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.97, \\
    %
    \mu_W^{*(\cdot,1)} = 34.01, \quad \sigma_W^{*(\cdot,1)} = 22.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.1 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.83, \\
    %
    \mu_W^{*(\cdot,2)} = 490.06, \quad \sigma_W^{*(\cdot,2)} = 502.56.
    %
    \mu_A^{*(\cdot,3)} = 0.2 s, \enspace \sigma_A^{*(\cdot,3)} = 0.3 s, \enspace \phi_A^{*(\cdot,3)} = 0.97, \\
    %
    \mu_W^{*(\cdot,3)} = 9154.16, \quad \sigma_W^{*(\cdot,3)} = 13538.75.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure 1 of supplement B shows one realization of $\Z$ for five dives of one simulated data set along with the corresponding reconstructed realization of $Y^*$.

The two simulated dive types differ in that dives of type 1 are much shorter on average (27 seconds) than dives of type 2 (128 seconds). The two simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn is much higher than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.

We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using a similar procedure as the one from the case study. Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory per data set.

For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states as follows.
Each test data set consists of a sequence of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
To assess the coarse-scale hidden state prediction, we estimate $p_t(i|y,\z) \equiv \Pr(X_t=i|Y=y,\Z=\z)$, $i=1,2$, $t=1,\ldots,100$ using the test-set observations $(y,\z)$ and training-set maximum likelihood estimates. These estimates are found using the {\em{forward-backward}} algorithm \citep{Zucchini:2016}. We compare these estimated conditional probabilities to $x_1,\ldots,x_{100}$, the true coarse-scale state realizations in the test data, by calculating the {\em{average dive decoding accuracy}} for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t|y,\z)/100$. We then report the average of these over the 500 training/test data set pairs. 
Analogously, to assess prediction of the fine-scale states, we estimate $p^*_{t,\tilde t^*}(i^*|y,\z) \equiv \Pr(\tilde X^*_{t,\tilde t^*}=i|Y=y,\Z=\z)$, $i^*=1,2,3$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, $t=1,\ldots,100$, using the test-set observations, the training-set maximum likelihood estimates, and the forward-backward algorithm. Denoting the true fine-scale state realizations from the test data set as $\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows. The conditional probabilities are estimated according to each of the four models under study using the maximum likelihood estimates from the training data set in conjunction with the forward-backward algorithm.

@@

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. Its average dive decoding accuracy is approximately $0.96$ and its average subdive decoding accuracy is approximately $0.91$. 
All parameter estimates of emission distributions ($\theta$ and $\theta^*$) and probability transition matrices ($\Gamma$,$\Gamma^*$) on both the coarse scale and fine scale are either comparable or favourable to all other models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors. This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to four times as large as the estimated standard error. See Tables 2 through 6 of supplement B for detailed results.
%All parameter estimates of fine-scale mean values ($\hat \mu^*$) and probability transition matrices ($\hat \Gamma^*$ and $\hat \Gamma$) have statistically insignificant biases. In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models.

The HHMM-DFT has a average dive decoding accuracy is comparable to the CarHHMM-DFT ($0.96$), but its average subdive decoding accuracy is worse by approximately 7 percentage points ($0.84$). Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$ and slightly overestimates $\sigma_A^{*(\cdot,3)}$. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \mu_A^{*(\cdot,23)}$, $\hat \sigma_A^{*(\cdot,1)}$, $\hat \sigma_A^{*(\cdot,2)}$ and $\hat \sigma_A^{*(\cdot,3)}$ are much smaller than the associated empirical standard errors (see Table 3 of supplement B). These results suggest that estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored. This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT. When standard errors are underestimated, the associated confidence intervals are too narrow, implying that researchers may be overconfident in their parameter estimates.

The CarHHMM is the worst-performing model in terms of dive decoding accuracy, which is approximately 4 percentage points worse than the CarHHMM-DFT ($0.92$). Its average subdive decoding accuracy is approximately 6 percentage points worse than the CarHHMM-DFT ($0.85$). This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct difference between the subdive states. In addition to its relatively poor average decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating emission parameters and probability transition matrices. Estimates associated with subdive states 2 and 3 ($\theta^{*(\cdot,2)}, , \hat \mu_A^{*(\cdot,3)}$) are especially poor. See supplement B for more detailed results.

Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than a third of that of the other models (see Table \ref{table:accuracy}). However, this model cannot estimate dive type as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 5 of supplement B).

>>>>

\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.97, \\
    %
    \mu_W^{*(\cdot,1)} = 34.01, \quad \sigma_W^{*(\cdot,1)} = 22.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.1 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.83, \\
    %
    \mu_W^{*(\cdot,2)} = 490.06, \quad \sigma_W^{*(\cdot,2)} = 502.56, \\
    %
    \mu_A^{*(\cdot,3)} = 0.2 s, \enspace \sigma_A^{*(\cdot,3)} = 0.3 s, \enspace \phi_A^{*(\cdot,3)} = 0.97, \\
    %
    \mu_W^{*(\cdot,3)} = 9154.16, \quad \sigma_W^{*(\cdot,3)} = 13538.75.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure 1 of supplement B shows one realization of $\Z$ for five dives of one simulated data set along with the corresponding reconstructed realization of $Y^*$.

The two simulated dive types differ in that dives of type 1 are much shorter on average (27 seconds) than dives of type 2 (128 seconds). The three simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn is much higher than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.

We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using a similar optimization procedure as the one from the case study. %Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory per data set.

For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states as follows.
Each test data set consists of a sequence of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
To assess the coarse-scale hidden state prediction, we estimate $p_t(i|y,\z) \equiv \Pr(X_t=i|Y=y,\Z=\z)$, $i=1,2$, $t=1,\ldots,100$ using the test-set observations $(y,\z)$ and training-set maximum likelihood estimates. These estimates are found using the {\em{forward-backward}} algorithm \citep{Zucchini:2016}. We compare these estimated conditional probabilities to $\{x_1,\ldots,x_{100}\}$, the true coarse-scale state realizations in the test data, by calculating the {\em{average dive decoding accuracy}} for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t|y,\z)/100$. We then report the average of these over the 500 training/test data set pairs. 
Analogously, to assess prediction of the fine-scale states, we estimate $p^*_{t,\tilde t^*}(i^*|y,\z) \equiv \Pr(\tilde X^*_{t,\tilde t^*}=i|Y=y,\Z=\z)$, $i^*=1,2,3$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, $t=1,\ldots,100$, using the test-set observations, the training-set maximum likelihood estimates, and the forward-backward algorithm. Denoting the true fine-scale state realizations from the test data set as $\{\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}\}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows. The conditional probabilities are estimated according to each of the four models under study using the maximum likelihood estimates from the training data set in conjunction with the forward-backward algorithm.

@@

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. Its average dive decoding accuracy is approximately $0.96$ and its average subdive decoding accuracy is approximately $0.91$. 
All parameter estimates of emission distributions ($\theta$ and $\theta^*$) and probability transition matrices ($\Gamma$,$\Gamma^*$) on both the coarse scale and fine scale are either comparable or favourable to all other models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors. This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to five times as large as the estimated standard error. See Tables 2 through 6 of supplement B for detailed results.
%All parameter estimates of fine-scale mean values ($\hat \mu^*$) and probability transition matrices ($\hat \Gamma^*$ and $\hat \Gamma$) have statistically insignificant biases. In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models.

The HHMM-DFT has a average dive decoding accuracy is comparable to the CarHHMM-DFT ($0.96$), but its average subdive decoding accuracy is worse by approximately 7 percentage points ($0.84$). Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$ and slightly overestimates $\sigma_A^{*(\cdot,3)}$. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \mu_A^{*(\cdot,3)}$, $\hat \sigma_A^{*(\cdot,1)}$, $\hat \sigma_A^{*(\cdot,2)}$ and $\hat \sigma_A^{*(\cdot,3)}$ are much smaller than the associated empirical standard errors (see Table 3 of supplement B). These results suggest that estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored. This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT. When standard errors are underestimated, the associated confidence intervals are too narrow, implying that researchers may be overconfident in their parameter estimates.

The CarHHMM is the worst-performing model in terms of dive decoding accuracy and is approximately 4 percentage points worse than the CarHHMM-DFT ($0.92$). Its average subdive decoding accuracy is approximately 6 percentage points worse than the CarHHMM-DFT ($0.85$). This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct difference between the subdive states. In addition to its relatively poor average decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating emission parameters and probability transition matrices. Estimates associated with subdive states 2 and 3 ($\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}$) are especially poor. See supplement B for more detailed results.

Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than a third of that of the other models (see Table \ref{table:accuracy}). However, this model cannot differentiate between dive types as it assumes that there is only one. The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives. The resulting parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$) are highly correlated (See Figure 5 of supplement B).
